---
title: "Sine Waves and Beating"
format: html
execute: 
  echo: true
  warning: false
  message: false
---


# Introduction

In this document, we will explore how simple sine waves can interact to produce **beating patterns** and how these can be visualised using a **continuous wavelet transform**.  This exercise is designed to show the phenomenon of **amplitude modulation** clearly.

Before starting, make sure you have installed the package `gtseries`, which contains useful functions for spectral analysis, including the wavelet transform. You can install it with:

```r
install.packages("devtools")
devtools::github_install("mcrucifix/tseries.git")
```

Then, load it:

```{r}
library(gtseries)
```


# Defining Frequencies

We first define a time axis and a few close frequencies. These will allow us to observe beating patterns.

```{r}
t <- seq(32768)

# Define frequencies
w1 <- 1/100
w2 <- 1/104
w3 <- 1/130
w4 <- 1/134
```


# Visualising Frequencies

Let us plot vertical bars to represent the different frequencies.

```{r}
plot(c(w1, w2, w3, w4), rep(1, 4), type = "h", lwd = 3,
     xlab = "Frequency", ylab = "Amplitude", main = "Frequencies")
```

👉 **Think:** Why do we expect interference (beating) when frequencies are close to each other?


# Constructing Sine Waves

We now build individual sine waves.

```{r}
s1 <- ts(sin(t*w1), start = 0, deltat = 1)
s2 <- ts(sin(t*w2), start = 0, deltat = 1)
s3 <- ts(sin(t*w3), start = 0, deltat = 1)
s4 <- ts(sin(t*w4), start = 0, deltat = 1)
```


# Beating Between $w_1$ and $w_2$

Let us first combine two close frequencies, $w_1$ and $w_2$.

```{r}
plot(s1 + s2,   type = 'l', main = "Beating between w1 and w2")
lines(2*cos(0.5*(w1-w2)*t), col = 'blue', lwd = 3)
lines(-2*cos(0.5*(w1-w2)*t), col = 'blue', lwd = 3)
```

👉 **Observe:** The envelope (in blue) shows the amplitude modulation resulting from interference.

We can also look at the wavelet transform:

```{r}
plot(cwt_morlet(s1 + s2))
```


# Beating Between $w_1$ and $w_3$

Now we try with frequencies that are farther apart.

```{r}
plot(s1 + s3, type = 'l', main = "Beating between w1 and w3")
lines(2*cos(0.5*(w1-w3)*t), col = 'blue', lwd = 3)
lines(-2*cos(0.5*(w1-w3)*t), col = 'blue', lwd = 3)
```

👉 **Question:** How does the modulation period compare with the case of $w_1$ and $w_2$?

```{r}
plot(cwt_morlet(s1 + s3))
```


# Combining Three Frequencies

Now let us add three frequencies together. This will create a more complex beating structure.

```{r}
sum3 <- s1 + s2 + s3

plot(sum3, type = 'l', ylim = c(-6,6), main = "Combination of three frequencies")
lines(2*cos(0.5*(w1-w3)*t), col = 'blue', lwd = 3)
lines(2*cos(0.5*(w2-w3)*t), col = 'blue', lwd = 3)
lines(2*cos(0.5*(w2-w3)*t) + 2*cos(0.5*(w2-w3)*t), col = 'green', lwd = 3)
lines(-2*cos(0.5*(w2-w3)*t) - 2*cos(0.5*(w2-w3)*t), col = 'green', lwd = 3)
lines(3*cos(0.5*(w1-w2)*t), col = 'red', lwd = 3)
lines(-3*cos(0.5*(w1-w2)*t), col = 'red', lwd = 3)
```

```{r}
plot(cwt_morlet(sum3))
```

👉 **Experiment:** Try adding the fourth frequency `w4`. How does the pattern become more complex?


# Conclusion

In this exercise, we have seen how **beating patterns** emerge from the superposition of close sine waves, and how the **wavelet transform** provides a time-frequency perspective of the modulation. This is a key concept for understanding **amplitude modulation** in cyclostratigraphy. 

