
---
title: "Calculating Eccentricity Periods"
format: html
self-contained: true
---



## Introduction

In this exercise, we will use a set of fundamental frequencies, often denoted as $g_i$, to calculate a simplified version of the orbital eccentricity, $e$. These frequencies are derived from celestial mechanics and describe the long-term behavior of planetary orbits.

```{r}
# Load the vector of g_terms from Laskar's 1988 article
g_terms <- c(5.5689, 7.4555, 17.3769, 17.9217, 4.2489, 27.9606, 3.0695, 0.6669)

# Create a nice table for the students to see the values
laskar_terms_table <- data.frame(
  Term = paste0("g", 1:8),
  Value = g_terms
)

# Display the table
knitr::kable(laskar_terms_table, caption = "Laskar's g_terms as defined in the 1988 article")
```


The formula we will use for this demonstration is a simplified form of a more complex expression from celestial mechanics:

$$e = \sqrt{ \sum (\sin (g_i t + \phi_i))^2 + \sum (\cos (g_i t + \phi_i))^2 }$$

We will break down the calculation into a few key steps to make it easy to follow.

```{r}

#| include: false

library(gtseries)

```

## Step 1: Unit Conversion

The frequencies, $g_i$, are typically given in **arcseconds per year ($''$/yr)**. For our calculations, we need to convert them to **radians per kiloyear (rad/kyr)** to be consistent with the time variable, $t$, which we'll use in kiloyears.

To do this, we need to multiply our initial values by a series of conversion factors.

* To convert arcseconds to radians, we use the fact that $2\pi$ radians is equal to $360 \times 60 \times 60$ arcseconds. So the conversion factor is $\frac{2\pi}{360 \times 60 \times 60}$.
* To convert years to kiloyears, we simply multiply by 1000.

Can you justify why these terms are necessary for the conversion?

```{r}
# Convert g_terms from arcseconds/year to radians/kiloyear
# Conversion factor: (2 * pi / (360 * 60 * 60)) * 1000
conversion_factor <- (2 * pi / (360 * 60 * 60)) * 1000

# Apply the conversion to create a new vector
g_rad_kyr <- g_terms * conversion_factor

# Display the new vector
g_rad_kyr

```

The formula includes a phase term, $\phi_i$. In a full calculation, these phases are crucial, but for this illustrative purpose, we can simplify our problem. We will assume all phase terms are zero, i.e., $\phi_i = 0$.


## Step 2: Calculating Eccentricity

For this calculation, we will use the first two frequency terms, $g_2$ and $g_5$. We will also choose an arbitrary time, $t$, in kiloyears, for our calculation. Let's use $t=8$ Myr.

```{r}
# Let's define our time in kiloyears
t <- seq(8192)

# Extract the first two frequencies and their corresponding phases

# Calculate the arguments for the sine functions
arguments1 <- (g_rad_kyr[2] * t) 
arguments2 <- (g_rad_kyr[5] * t)

# Calculate the components for the formula. Note the parentheses correction.
esinPi <- (sin(arguments1) + sin(arguments2))
ecosPi <- (cos(arguments1) + cos(arguments2))


# Apply the full formula with the two sums
eccentricity_calc <- sqrt(esinPi^2 + ecosPi^2)

# we make it as a timeseries object 
eccentricity_calc_ts <- ts(eccentricity_calc, deltat=1)

# Display the final calculated eccentricity
plot(eccentricity_calc_ts)

```

## Step 3: Spectral Analysis with gtseries

Now that we have our time series for eccentricity, we can use the `gtseries` package to look at its spectral content. This will show us which frequencies contribute to the overall signal.

First, let's look at the spectrum of the `eccentricity_calc_ts` time series. The `mfft_real` function helps us find the discrete spectrum.

```{r}
require(gtseries)
# Calculate the real spectrum of the eccentricity time series
real_spectrum <- mfft_real(eccentricity_calc_ts)

# Plot the spectrum to see the dominant frequencies
print(real_spectrum)
plot(real_spectrum)

```

Notice that the spectrum of the full eccentricity signal does **not** simply show the initial $g_1$ and $g_2$ frequencies. Because the eccentricity is a non-linear combination of these terms (specifically, it involves squares and a square root), the resulting spectrum shows new, prominent frequencies. The most dominant peaks correspond to the sum and difference frequencies of the initial terms, such as $g_3-g_4$. This is fundamental... and not surprising. Think that eccentricity is the _enveloppe_ of the precession signal ($e\sin\Pi$). As we have seen in the theory of beatings, it will only include differences (and because this enveloppe is here rectified, it also contains many harmonic, higher-order terms)

## Step 4: Combining Frequencies and Exploring Hierarchy

We've seen how combining just two frequencies creates a beat signal. Now, let's explore this further by combining more frequencies. Based on our exercises on beatings, we know that the combination of different sine waves will produce a complex signal with new frequencies corresponding to the differences of the original frequencies.

This exercise will show the hierarchy of the 100-ka, 400-ka, and 2.4-Ma cycles in Earth's eccentricity. These are some of the most important periods in cyclostratigraphy analysis. 

Your challenge is to:

1.  **Add more terms** to the eccentricity calculation in **Step 3**. Try adding a third term, perhaps `g_rad_kyr[3]`, and see how the resulting spectrum changes.
2.  **Identify the new beat frequencies** that appear in the spectrum. You should be able to spot the frequencies corresponding to the 100-ka and 400-ka periods.
3.  **Think about the hierarchy.** What is the relatioship that we expect between the phases of the 100-ka, 400-ka and 2.4 Ma ? One key aspect that I would like you to see, is that the 100-ka cycles _vanishes_ near 2.4 Ma nodes. Can you show this ? 


To do this, I give you a help with a little helper function: 

```{r}
get_period <- function(one, two) {2*pi/abs( g_rad_kyr[one]  - g_rad_kyr[two])}
```

but it won't be enough. I also need to tell you a little secret: in the spectrum of eccentricity, you can also get third-order and even fourth-order combinations. And the other little secret: the so-called "404-ka" signal is actually made of two components: 404-ka and a weaker 486-ka component. 
